Explaining probabilistic prediction on the simplex with Shapley compositions

### 2023-12-20 paper/main.tex
edit paper and code

The projection of a composition into an element of the Aitchison orthonormal basis, i.e. a component of the ILR space. This can be interpreted as a \emph{balance}, i.e. a logratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}. One giving the central values of the parts in one group of parts and one for another group of parts. Therefore, in our context, a balances is comparing the weight of two group of classes. This can be illustrated by a sequential binary partition of bifurcation tree. Figure \ref{fig:trees} gives two examples.
\begin{figure}
  \centering
  \subfigure[]{\label{fig:bifurc1}
    \input{bifurctree.tex}
    }
  \subfigure[]{\label{fig:bifurc2}
    \input{bifurctree.tex}
  }
  \caption{Two examples of bifurcation tree. \ref{fig:bifurc1} is the one corresponding to the basis obtained by with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}...}
  \label{fig:trees}
\end{figure}

@@

Some may find the fine analysis of the features contributions in cases with more than four classes tricky. Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:moreclasses}, choosing which subspaces to visualize require a careful understanding of the sequential binary partition. However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections. Indeed the norm of a Shapley composition gives the strength of the feature's contribution in the prediction. It gives the overall contribution of the feature on the prediction, regardless of its direction. The angle between two Shapley compositions can informs about their orthogonality. If two Shapley composition are orthogonal\footnote{We refer here to geometric orthogonality, not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}, this suggest the features are nonredundant. A negative angle would suggest that the features have an opposite influence on the prediction. The projection of a Shapley composition on the set of class-compositions informs in favor of, or against, which classes a feature is contributing. Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation using the norm of Shapley compositions, angles between them and their projection on the class-compositions.

>>>>

A component of the ILR space  can be interpreted as a \emph{balance}, i.e. a logratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}: one giving the central values of the probabilities in one group of classes and one for another group of classes. Therefore, a balance is here comparing the weight of two group of classes. This can be illustrated by a sequential binary partition or bifurcation tree. Figure \ref{fig:trees} gives two examples. \ref{fig:bifurc1} shows the bifurcation tree corresponds to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$. The first balance $\tilde{p}_1$ first compare the probabilities of class $1$ and $2$. Each next balance then recursively compares the probability for the next classe with the probabilities for the previous ones. \ref{fig:bifurc2} shows another bifurcation tree....
\begin{figure}
  \centering
  \subfigure[]{\label{fig:bifurc1}
    \input{bifurctree.tex}
    }
  \subfigure[]{\label{fig:bifurc2}
    \input{bifurctree1.tex}
  }
  \caption{Two examples of bifurcation tree. \ref{fig:bifurc1} corresponds to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum} is another example corresponding to the basis used in the example...}
  \label{fig:trees}
\end{figure}

@@

Some may find the fine analysis of the features contributions in cases with more than four classes tricky. Indeed, in this case, the ILR space cannot be visualized in a $2$ or $3$-dimensional plot and as discussed in Section \ref{sec:balances}, choosing which subspaces to visualize require a careful understanding of the sequential binary partition. However, as we already had the insight from the above visualization, the Shapley explaination can be summarized by sets of angles, norms and projections. Indeed the norm of a Shapley composition gives the strength of the feature's contribution in the prediction. It gives the overall contribution of the feature on the prediction, regardless of its direction. The angle between two Shapley compositions can informs about their orthogonality. If two Shapley composition are orthogonal\footnote{We refer here to geometric orthogonality, not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}, this suggest the features are nonredundant. A negative angle would suggest that the features have an opposite influence on the prediction. The projection of a Shapley composition on the set of class-compositions informs in favor of, or against, which classes a feature is contributing. Appendix \ref{app:summarize}, provides some examples of summarizing a Shapley explanation using the norm of Shapley compositions, angles between them and their projection on the class-compositions.
